---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data
```{r,results='asis',echo=TRUE,fig.height=3,fig.width=5}
Sys.setlocale("LC_TIME", "C")
dataFileNAme="activity.zip"
unzip(dataFileNAme)
reprodata <- read.csv("activity.csv",header = TRUE)
totalStepPerDay <- sapply(split(reprodata,reprodata$date),function(x) sum(x$steps,na.rm=TRUE))
medianStepPerDay <- median(totalStepPerDay,na.rm=TRUE)
meanStepPerDay <- mean(totalStepPerDay,na.rm=TRUE)

vmedian_text=paste("median: ",medianStepPerDay)
vmean_text=paste("mean: ",meanStepPerDay)
library(reshape2)

reprodataReshaped=melt(reprodata[,c("interval","steps")],id="interval", measure.vars=c("steps"),variable.name="steps", na.rm=TRUE)
reprodataMean= dcast(reprodataReshaped, interval~ steps,mean)
 

require(ggplot2)
require(plyr)


```





## What is mean total number of steps taken per day?

1. Make a histogram of the total number of steps taken each day

```{r}
hist(totalStepPerDay,col="light green",main="Histogram Total number of steps taken per day", xlab="steps")
abline(v=meanStepPerDay,col="red")
abline(v=medianStepPerDay,col="blue")
legend("topright", cex=0.5,col=c("red","blue"), lty=c(1,1), legend=c(vmedian_text,vmean_text))
 

```


2. Calculate and report the mean and median total number of steps taken per day


  Median is:
```{r,echo=TRUE,results='asis'}
 median(totalStepPerDay,na.rm=TRUE)
```

Mean is: 
```{r}
mean(totalStepPerDay,na.rm=TRUE)

```



## What is the average daily activity pattern?

1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r}
ggplot(reprodataMean, aes(interval,steps)) + geom_line(color="blue")+ labs(title="steps average daily activity")+ labs(x="5-minute interval")+ labs(y="averaged across all days")  
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

    Interval with maximal number of steps: 


```{r}
max <- reprodataMean[reprodataMean$steps ==max(reprodataMean$steps),]

max$interval[1]
 
```


## Imputing missing values

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r}
numberOfNa <- length(which((is.na(reprodata))))
numberOfNa
```

2.  Strategy for filling in all of the missing values in the dataset:

    - The missing value will be replaced with the mean for that 5-minute interval.

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}
reprodataFilled <- reprodata
numRow <- nrow(reprodataFilled)
j =1
numberOfReplacedNa=0
for(i in 1:numRow){
   
  if(is.na(reprodataFilled$steps[i])){
    reprodataFilled$steps[i] = reprodataMean$steps[j]
     numberOfReplacedNa=numberOfReplacedNa+1
  }
  j=j+1
  if(j==289){
    j=1
  }
}
head(reprodataFilled,n=10)
```

4.     Make a histogram of the total number of steps taken each day
```{r}
reprodataFilledReshaped <- melt(reprodataFilled[,c("interval","steps")],id="interval", measure.vars=c("steps"),variable.name="steps", na.rm=FALSE)
reprodataFilledSum= dcast(reprodataFilledReshaped, interval~ steps,sum)
meanStepPerDayFilled <-mean(reprodataFilledSum$steps)
medianStepPerDayFilled <- median(reprodataFilledSum$steps)
hist(reprodataFilledSum$steps,col="light blue",main="Histogram Total number of steps per day approximated NAs", xlab="steps")
abline(v=meanStepPerDayFilled,col="red")
abline(v=medianStepPerDayFilled,col="blue")
legend("topright", cex=0.4,col=c("red","blue"), lty=c(1,1), legend=c(vmedian_text,vmean_text))


```


- Calculate and report the mean and median total number of steps taken per day.
Mean is :
```{r}
meanStepPerDayFilled
```

Median is: 
```{r}
medianStepPerDayFilled
```



- Do these values differ from the estimates from the first part of the assignment? 

The values differ from those estimated in the first part
```{r }
meanStepPerDayFilled < meanStepPerDay && medianStepPerDayFilled < medianStepPerDay 
```

- What is the impact of imputing missing data on the estimates of the total daily number of steps?

    Removing the NAs biases the results. Mean and median appear to be greater than they are.

## Are there differences in activity patterns between weekdays and weekends?

 - Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

```{r}
weekDays<-c("Monday","Tuesday","Wednesday","Thursday","Friday")
#weekEnd <-c("Saturday","Sunday")
 
reprodataFilled <-transform(reprodataFilled,daytype = factor( x=ifelse(weekdays(as.Date(reprodataFilled$date)) %in% weekDays,"weekday","weekend"),levels = c("weekday","weekend") ))
summary(reprodataFilled)
head(reprodataFilled)
```


  
- Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). The plot should look something like the following, which was creating using simulated data:
We first melt and recast the data in order to bring it into a form that suitable for plotting:
```{r}
copyFilled <- reprodataFilled
copymelt <-  melt(copyFilled ,id=c("interval","date"), measure.vars=c("steps"),variable.name="steps", na.rm=TRUE)

copymeltMutate<-mutate(copymelt,daytype = factor( x=ifelse(weekdays(as.Date(copymelt$date)) %in% weekDays,"weekday","weekend"),levels = c("weekday","weekend") ))

copymeltMean= dcast(copymeltMutate, interval+daytype ~ steps,mean)



```

and then make the panel plot:
```{r}
#  qplot(interval,value,data=reprodataFilledReshaped,facets= daytype ~ .,geom =c("line","smooth"),color=daytype)
# require(lattice)
# xyplot(reprodataFilledReshaped$value ~ reprodataFilledReshaped$interval | reprodataFilledReshaped$daytype, layout=c(2,1))


ggplot(copymeltMean, aes(interval,steps)) + geom_line(aes(color = daytype))+geom_smooth(method="auto")+facet_grid(. ~ daytype)+ labs(title="steps average daily activity")+ labs(x="5-minute interval")+ labs(y="averaged across all days")  


```

